Non-word repetition in Yélî Dnye

### 2020-09-09 manuscript.Rmd
add more phon info to intro

### Intro to the language (**mc**) - please feel free to throw away anything that is not useful!

- complexity in the vowel system
- complexity in the consonant system
- word shapes
- typical word length
- although not the focus of this paper, high use of suppletion in verbal paradigms, other features of language, see Levinson XXX for details

@@

>>>>

### Yélî Dnye phonological overview
<!--M2A: is this okay?-->
Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7000 people residing on Rossel Island, which is found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea. The Yélî sound system, much like its baroque grammatical system [@levinsoninprepydgrammar], is unlike any other in the region.

With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments. The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of these able to be contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/). A large number of this combinatorial set can further be contrastively labialized or palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see @levinsoninprepydgrammar for details). The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in Yélî Dnye may be oral or nasal, short or long. The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels, with 7 of these able to appear as short and long nasal vowels as well (/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).

In total, Yélî Dnye uses 90 distinctive segments (93 when including consonants that are extremely rarely used), far outstripping the phonemic inventory size of other documented Papuan languages (@levinsoninprepydgrammar; Foley 1986) and includes at least two contrasts not yet been documented elsewhere [labial-coronal double-articulations with dental vs. post-alveolar coronal placement in both oral and nasal stops, Maddieson & Levinson 2001; Ladefoged and Maddieson 1996].

Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllablic words appearing least frequently [~10%; based on >5800 lexemes in the current Yélî Dnye dictionary, @levinsoninprepydgrammar]. The vast majority of syllables use a CV format. A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., 'ndap' (Spondylus shell) /ṇṭæp/). However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CVCV structure instead (e.g., 'ndapî' /'ṇṭæ.pɯ/). This process is used frequently with English loan words that have a coda (e.g., 'ponî' (phone) /'pɔ.nɯ/). An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but these are limited to /æ/-. Finally, the lexicon features a handful of single-vowel grammatical morphemes (see @levinsoninprepydgrammar for details).

While all consonants can appear in the first syllable of the word, in bisyllabic roots, the second syllable onset consonant is almost exclusively drawn from a subset of the simplex consonants (in order of descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/, /t/, /ɣ/, /mʷ/). In monosyllabic stems, oral vowels (short: 43% and long: 31%) are more common than nasal ones (short: 4% and long: 22%), with the first syllable of bisyllabic stems following a similar pattern. Most vowels in the second syllable of bisyllabic stems are also short and oral, with a potential preference for rounding harmony between the first and second vowels. Stress typically falls on the first syllable of bisyllabic words, on the first and third syllables of tetrasyllabic words, and on the first or second syllable of trisyllabic words, depending on its vocalic features (see @levinsoninprepydgrammar for details).

@@

Most speakers of the language grow up speaking it monolingually within their swidden horticulturalist communities until they begin attending school around the age of 7; school instruction is in English. While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (1953--present) where there appears to be a higher incidence of married-in mothers from other islands [@brownIPchapter], the same region in which the current data were collected. Children in these multilingual families grow up speaking Yélî Dnye and English, Tok Pisin, and/or some other language(s) from the region.
